---
title: "Inventory size, similarity and dissimilarity"
author: "Carlos Silva and Steven Moran"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
    toc: true
---

# Introduction

Load packages

```{r}
library(tidyverse)
library(ggplot2)
library(ggrepel)
```


Get data

```{r}
inv <- read.csv("Inventories.csv")
head(inv)
```

Prepare data

```{r}
df_inv_long <- inv %>% select(Language, Phoneme) %>%  mutate(newcol = 1)

df_inv <- df_inv_long %>% pivot_wider(names_from = Language, values_from = newcol, values_fill = 0)

df_total_inv <- df_inv %>% mutate_at(c(2:38), as.numeric)

head(df_total_inv)

```

# Inventory size

Get the consonant inventory size for all languages

```{r}

cons_count <- df_total_inv %>% select(c(2:38)) %>% mutate_at(c(1:37), as.numeric)
  
count <- colSums(cons_count [,c(1:37)]) %>% unname(colSums(count))

cons_lg <- select(df_inv_long, 'Language') 

Language <- unique(cons_lg$Language)

category <- inv %>% select(Language, Category)

count_lg <- data.frame(cbind(Language, count)) 
  
count_lg_1 <- inner_join(count_lg, category, by = "Language") %>% distinct()

inv_size <- transform(count_lg_1, count = as.numeric(count))

head(inv_size)

```

Which languages have bigger inventories?

```{r}
ggplot(inv_size) + 
  geom_bar(aes(x = Language,
               y = count,
              fill = Category), 
           stat = "identity", 
           show.legend = TRUE,
           position = "dodge2") + coord_flip()
```


Violin plot: the majority of creoles have larger consonant inventories than Portuguese.

```{r}
ggplot(inv_size, aes(x = Category, y = count, fill = Category)) +
  geom_violin(alpha = 0.5) +
  geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5) +
  theme(legend.position = "none")
```




Is there a relationship between the inventory size of creoles and the inventory size of substrates?

NO ANSWER YET


# Consonant frequency in all languages involved


Count frequent consonants

```{r}
total <- rowSums(cons_count)

cons_freq <- data.frame(cbind(df_total_inv$Phoneme, total))

cons_freq <- transform(cons_freq, total = as.numeric(total))

colnames(cons_freq)[1] <- "LexifierPhoneme"

```




## Is there a relationship between this frequency and the stability values?

Subset: Portuguese consonants only
```{r}
cons_freq_pt <- cons_freq %>% subset(LexifierPhoneme %in% c('b','d','f','g','k','l','ʎ','m','n','ɲ','p','t̠ʃ','ɾ','s',
                                       't','v','z','ʒ', 'r'))

```

First dataset: relative frequency values
```{r}
cons_freq_rel <- cons_freq_pt %>% mutate(frequency = total/37)
```

Second dataset:stability values  
```{r}
consonant_global_stability <- read.csv("consonant_global_stability.csv")
```

Merge datasets
```{r}
cor_freq_sta <- left_join(consonant_global_stability, cons_freq_rel, by='LexifierPhoneme')
```

Results of a simple regression
```{r}
fs <- lm(frequency ~ mglobal, data=cor_freq_sta)
summary(fs)
```

Plot the results
```{r}

fs_plot <- ggplot(cor_freq_sta, aes(x = frequency, y = mglobal, label = LexifierPhoneme)) +
  geom_smooth(method = "lm") +
  geom_point() # +
  #geom_text(aes(label=V1), hjust=3, vjust=0)

fs_plot + geom_text_repel(aes(label=LexifierPhoneme))
```

There relationship between stability and frequency across all languages involved. But does it make sense? We are measuring the consonants in all categories and the present of the lexifier and the creoles may influence the results. Perhaps, we should try to subset and measure the frequency across substrates only, but I think that this procedure would just increase the lack of correlation.

## Is there a relationship between consonant stability and their presence in the inventories of the substrate languages?


Data preparation (substrates only)

```{r}
inv_subs <- inv %>% subset(Category == 'substrate') %>%  select(Language, Phoneme)  %>% mutate(newcol = 1)

inv_subs_long <- inv_subs %>% pivot_wider(names_from = Language, values_from = newcol, values_fill = 0)

inv_subs_long <- inv_subs_long %>% mutate_at(c(2:18), as.numeric)

head(inv_subs_long)

```

Sum row values and subset to consonants which have correspondents in Portuguese

```{r}

subs_count <- inv_subs_long %>% select(c(2:18))

total_subs <- rowSums(subs_count)

subs_freq <- data.frame(cbind(inv_subs_long$Phoneme, total_subs))

subs_freq <- transform(subs_freq, total_subs = as.numeric(total_subs))

colnames(subs_freq)[1] <- "LexifierPhoneme"

subs_freq <- subs_freq %>% subset(LexifierPhoneme %in% c('b','d','f','g','k','l','ʎ','m','n','ɲ','p','t̠ʃ','ɾ','s', 't','v','z','ʒ', 'r'))

```

Get relative values

```{r}
subs_freq_rel <- subs_freq %>% mutate(frequency = total_subs/17)
```

Merge datasets
```{r}
subs_sta <- left_join(consonant_global_stability, subs_freq_rel, by='LexifierPhoneme')
```

Results of a simple regression
```{r}
subs_sta_lm <- lm(frequency ~ mglobal, data=subs_sta)
summary(subs_sta_lm)
```
Plot the results
```{r}
subsfreq_sta_cor <- ggplot(subs_sta, aes(x = frequency, y = mglobal, label = LexifierPhoneme)) +
  geom_smooth(method = "lm") +
  geom_point() +
  xlab("Frequency across substrates") + ylab("Stability across creoles") # +
 # geom_text(aes(label=V1), hjust=3, vjust=0)


subsfreq_sta_cor + geom_text_repel(aes(label=LexifierPhoneme))
```

When we consider the correlation between the consonant stability in creoles and the frequency of these consonants in the substrates only, we find that there is a weaker correlation (if we compare with the results above). However, this correlation is statistically significant (p-value: 0.004). Nothing the outliers which are normally voiced consonants.

# Typological frequency vs. Substrate frequency

Typological frequency data
```{r}
typ_freq <- read.csv("typ_freq.csv")
```

Merge datasets

```{r}
#consonant_global_stability <- read.csv("consonant_global_stability.csv")

typ_sta <- left_join(typ_freq, consonant_global_stability, by='LexifierPhoneme')
```

Results of a simple regression
```{r}
typ_sta_lm <- lm(TypologicalFreq ~ mglobal, data=typ_sta)
summary(typ_sta_lm)
```
Plot the results
```{r}
typ_sta_cor <- ggplot(typ_sta, aes(x = TypologicalFreq, y = mglobal, label = LexifierPhoneme)) +
  geom_smooth(method = "lm") +
  geom_point() +
  xlab("Frequency across world's languages") + ylab("Stability across creoles") # +
 # geom_text(aes(label=V1), hjust=3, vjust=0)


typ_sta_cor + geom_text_repel(aes(label=LexifierPhoneme))

```




# Inventory similarity and overall creole stability

Getting just the consonants which have correspondents in Portuguese

```{r}
inv_overlap <- df_total_inv %>% subset(Phoneme %in% c('b','d','f','g','k','l','ʎ','m','n','ɲ','p','ʀ','ɾ','s','ʃ',
                                       't','v','z','ʒ'))

```

Which languages have more have more consonants that overlap with Portuguese consonants

```{r}
cons_over <- inv_overlap %>% select(c(2:38)) %>% mutate_at(c(1:37), as.numeric)
  
absolute_overlap <- colSums(cons_over [,c(1:37)]) %>% unname(colSums(count))

cons_overlap <- select(df_inv_long, 'Language') 

Language <- unique(cons_lg$Language)

category <- inv %>% select(Language, Category)

overlap <- data.frame(cbind(Language, absolute_overlap)) 
  
inv_overlap <- inner_join(overlap, category, by = "Language") %>% distinct()

head(inv_overlap)
```

Adding a new variable: inventory dissimilarity

```{r}
colnames(inv_overlap)[2]  <- "similar_consonants"
colnames(inv_size)[2] <- "total_consonants"


df_st <- merge(inv_overlap, inv_size, by = "Language")

df_st$total_consonants = as.numeric(as.character(df_st$total_consonants))
df_st$similar_consonants = as.numeric(as.character(df_st$similar_consonants))


df_st$different_consonants <- df_st$`total_consonants` - df_st$`similar_consonants`
```

Transforming values from absolute to relative

```{r}
tsd_rel <- df_st %>%  mutate(total_rel = total_consonants/19, similar_rel = similar_consonants/19, diff_rel= different_consonants/19)
```

### Does substrate similarity influence creole stability?

If yes, we expect to find a positive correlation overall creole stability ~ substrate inventory similarity, and a negative correlation overall creole stability ~ substrate inventory divergence.

Let's just subset the data to substrates only.

```{r}
subs_tsd_rel <- tsd_rel %>% subset(Category.x=='substrate')
```

Prepare data to get the correlation between creole stability and substrate similarity

```{r}

Language <- c('Angolar',
'Cape Verdean Brava',
'Cape Verdean Fogo',
'Cape Verdean Santiago',
'Cape Verdean Santo Antão',
'Cape Verdean São Vicente',
'Daman',
'Diu',
'Fa dAmbô',
'Guinea-Bissau Kriyol',
'Kannur',
'Korlai',
'Papiá Kristang',
'Patua Macau',
'Principense',
'Santome',
'Sri Lanka',
'Timor creole',
'Casamancese')

value <- c(0.6571429	,
0.9285714	,
0.8857143	,
0.9189189	,
0.9444444	,
0.9142857	,
0.9000000	,
0.8857143	,
0.6428571	,
0.9054054	,
0.9242424	,
0.8529412	,
0.8750000	,
0.9054054	,
0.8243243	,
0.7702703	,
0.8970588	,
0.9000000	,
0.8750000)

creole_stability_simple <- data.frame(Language, value)

```

```{r}

substrate_similar_simple <- subs_tsd_rel %>%  select(Language, similar_rel)

colnames(substrate_similar_simple)[2]  <- "value"

```


Let's import a new dataframe to get language pairs

```{r}
correspondence <- read.csv("substrate_creole_correspondence.csv")
```

Join stability and similarity values to the new data frame

```{r}
colnames(creole_stability_simple)[1]  <- "creole"
colnames(creole_stability_simple)[2]  <- "stability"

cor1 <- left_join(correspondence, creole_stability_simple, by= 'creole')

colnames(substrate_similar_simple)[1]  <- "substrate"
colnames(substrate_similar_simple)[2]  <- "similarity"

cor2 <- left_join(cor1, substrate_similar_simple, by= 'substrate')
```

Results of a simple regression
```{r}
cor2_lm <- lm(similarity ~ stability, data=cor2)
summary(cor2_lm)
```
Plot the results
```{r}
ggplot(cor2, aes(x = similarity, y = stability, label = pair)) +
  geom_smooth(method = "glm") +
  geom_point() +
  geom_text(aes(label=pair), hjust=1, vjust=1) +
  facet_wrap(~ area)
```


It seems that stability increases as substrate similarity increases in the Indian creoles, but not in the other creoles, unless I am not getting this right. It would be nice to have the pairs displayed as labels, but I think that I would need to turn this graph into five different graphs.

# Jaccard distance

Jaccard between two vectors manually selected

```{r}
jaccard <- function(a, b) {
    intersection = length(intersect(a, b))
    union = length(a) + length(b) - intersection
    return (intersection/union)
}

#data <- df_total_inv

jaccard(df_total_inv$`Portuguese`, df_total_inv$`Cantonese`)


```




VEGAN PACKAGE -- Jaccard distance between all vectors at the same time

```{r}
#install.packages("vegan")
library(vegan)
df_jaccard <- select(df_total_inv, c(2:35))
results_jaccard <- vegdist(df_jaccard, method = "jaccard")
# I may have worked, but we still need a visualization.
```
ADE4 PACKAGE -- Same as above, but with hierarchical clustering-- still no visualization. Also note that the results of this vector are different from the results generated using the Vegan package

```{r}
#install.packages("ade4")
library(ade4)
mm1 <- as.matrix(df_jaccard)
mm2 <- matrix(mm1, ncol = ncol(df_jaccard), dimnames = NULL)
ID <- colnames(df_jaccard)
d <- dist.binary(mm2, method = 1, diag = FALSE, upper = FALSE) #method 1 is Jaccard index (1901) S3 coefficient of Gower & Legendre
hc <- hclust(d)               # apply hierarchical clustering 

```
Trying again, but just with the Jaccard function applied to languages in different cycles

```{r}
jaccard <- function(a, b) {
    intersection = length(intersect(a, b))
    union = length(a) + length(b) - intersection
    return (intersection/union)
}


vec_port <- df_total_inv[2]
vec_jac <- c() 
vec_crio_name <- c() 


#for (i in 2:35)
#{jaq_val <- jaccard(vec_port,df_total_inv[i])
#	vec_jac <- c(vec_jac, jaq_val)
#	vec_crio_name <- c(vec_crio_name, colnames(df_total_inv)[i])}



#jac_data <- data.frame(vec_jac, vec_crio_name)

```

##Steven's code

```{r}
library(tidyverse)

# Create Jaccard table based on full set of phonemes across all languages 
# in the sample; other ways to do this -- to discuss
df <- read_csv("Inventories.csv")
df <- df %>% select(Language, Phoneme)
df$presence <- 1
df_wide <- df %>% spread(Phoneme, presence)
df_wide <- df_wide %>% replace(is.na(.), 0)
head(df_wide)

# Corresponding languages for which we want Jaccard measurements
correspondences <- read_csv("substrate_creole_correspondence.csv")
tmp <- correspondences %>% slice(1:2)

# DONE: fix the language names in the correspondences table so they can be used across tables
all(correspondences$substrate %in% df_wide$Language) # OK
all(correspondences$creole %in% df_wide$Language) # OK
correspondences$substrate %in% df_wide$Language # OK
correspondences$creole %in% df_wide$Language # OK

# Get method for Jaccard
#library(qvalue) #Error in library(qvalue) : there is no package called ‘qvalue’
#library(jaccard) #Error: package ‘jaccard’ was built under R version 4.3.1Error: package or namespace load failed for ‘jaccard’ 

# Test 
 x <- as.numeric(df_wide[1,2:116])
 y <- as.numeric(df_wide[1,2:116])
 jaccard(x,y)
 x==y # All TRUE -- passed!

# Run Jaccard on test data -- some bug here in the code because new_col is the same
# tmp %>% mutate(new_col = jaccard(as.numeric(as.factor(df_wide %>% filter(Language == tmp$substrate) %>% select(2:116))), as.numeric(as.factor(df_wide %>% filter(Language == tmp$creole) %>% select(2:116)))))

# When all language names match in both tables this *should* run fine
# correspondences %>% mutate(new_col = jaccard(as.numeric(as.factor(df_wide %>% filter(Language == correspondences$substrate) %>% select(2:116))), as.numeric(as.factor(df_wide %>% filter(Language == correspondences$creole) %>% select(2:116)))))
```