---
title: "Supplementary materials for: Consonant stability in Portuguese-based creoles"
author: "Steven Moran and Carlos Silva"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
    toc: true
bibliography: 'bibliography.bib'
---
  
# Overview
  
Supplementary materials for [Consonant Stability in Portuguese-based creoles](https://www.overleaf.com/project/60cdac0dd5871295e0f608fc). Silva, Carlos and Steven Moran. Work in progress.

In this report, we use R [@R] and the following packages [@tidyverse;@knitr].

```{r, message=FALSE}
library(tidyverse)
library(knitr)
```

Load the dataset.

```{r, message=FALSE}
# Token keep resetting...
# database <- read_csv('https://raw.githubusercontent.com/CreoPhon/CreoPhonPt/main/Creoles.csv?token=AAIGDLUMP3R6LQG6Y6GJA6TBVDYJQ')
database <- read_csv('database.csv')
```

The data look like this.

```{r}
database %>% head() %>% kable()
```

Let's extend the database with some variables. Duration of contact.

```{r}
database$duration <- database$`EndOfInfluence` - database$`FirstMajorSettlement`
```

Next, a variable of global stability.

```{r}
database <- mutate(database, GlobalStability = (PlaceStability + MannerStability)/2)
# table(database$Language, database$GlobalStability)
```

Categorical variable for duration.

```{r}
database <- database %>% mutate(duration_group = ifelse(duration <= 200, 'short', 'long'))
```

Categorical variable for changes in manner and/or place. Stability in the database is '1' (no change) and '0' change.

```{r}
database <- database %>% mutate(categorical_stability = ifelse(PlaceStability == 1 & MannerStability == 1, 'no manner/no place', NA))

database <- database %>% mutate(categorical_stability = ifelse(PlaceStability == 1 & MannerStability == 0, 'manner/no place', categorical_stability))

database <- database %>% mutate(categorical_stability = ifelse(PlaceStability == 0 & MannerStability == 1, 'no manner/place', categorical_stability))

database <- database %>% mutate(categorical_stability = ifelse(PlaceStability == 0 & MannerStability == 0, 'manner/place', categorical_stability))

table(database$categorical_stability)
```


# Creole stability

Which creoles in the sample are more or less stable overall?

```{r, message=FALSE}
creole_stability <- database %>% group_by(Language, Area, duration, duration_group) %>% summarize(MeanStability = mean(GlobalStability, na.rm = TRUE))
```

Plot it.

```{r}
ggplot(creole_stability) + 
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), fill = Area), 
           stat = "identity", show.legend = TRUE) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score")
```

We have the overall stability values. What are these in relation to the duration of contact?

There does not seem to be a relationship between overall duration and overall stability.

```{r}
ggplot(creole_stability, aes(x=duration, y=MeanStability)) +
  geom_point()

ggplot(creole_stability, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text(label=creole_stability$Language)
```
Results from the simple regression.

```{r}
msd <- lm(MeanStability ~ duration, data=creole_stability)
summary(msd)
```

However, there does seem to be two groups of languages -- ones that belong to "long duration" (>= 400 years) and those that belonw to "short duration" (<= 200 years) (Nicholas Lester, pc).

We can try to split the data and rerun the models, but we note that there are very few data points.

```{r}
tmp_short <- creole_stability %>% filter(duration <= 200)
tmp_long <- creole_stability %>% filter(duration > 200)
```

```{r}
ggplot(tmp_short, aes(x=duration, y=MeanStability)) +
  geom_point()

ggplot(tmp_short, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text(label=tmp_short$Language)
```

```{r}
ggplot(tmp_long, aes(x=duration, y=MeanStability)) +
  geom_point()

ggplot(tmp_long, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text(label=tmp_long$Language)
```
Or perhaps a single model with an interaction term MeanSim ~ duration, group * duration.

```{r}
msd <- lm(MeanStability ~ duration + duration_group * duration, data=creole_stability)
summary(msd)

ggplot(creole_stability, aes(x = duration, y = MeanStability, color = duration_group)) +
  geom_smooth(method = "lm") +
  geom_point()
```

Nicholas Lester notes: The variability in the two groups is very different. The direction of the effect is interesting: shorter durations yield more stability more consistently. Over time, the variability in mean stability increases. Time is "destabillizing the pattern of stability".

And we can also increase the number of observations by running the analysis at the segment level, rather than on mean stability.


# Segment stability

Which segments are the most stable across creoles in the language sample?

We calculate stability of place and manner for each phoneme.

```{r}
place_results <- database %>% group_by(LexifierPhoneme) %>% summarize(mplace = mean(PlaceStability, na.rm = TRUE))
manner_results <- database %>% group_by(LexifierPhoneme) %>% summarize(mmanner = mean(MannerStability, na.rm = TRUE))

consonant_stability <- left_join(place_results, manner_results, by = "LexifierPhoneme")

class <- c("nasal", "rhotic", "lateral", "fricative", "stop", "stop", "fricative", "stop", "stop", "lateral", "nasal", "nasal", "stop", "rhotic", "fricative", "stop", "affricate", "fricative", "fricative")

consonant_stability_class <- cbind(consonant_stability, class)
```

Next, we plot the results.

```{r}
# TODO: fix the class "dots"
ggplot(consonant_stability, aes(y = mmanner, x = mplace, label = LexifierPhoneme, color=class)) +
  geom_point(position= "dodge") + 
  geom_text(aes(label=LexifierPhoneme), hjust=3, vjust=0) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

Here is an alternative view for the global results.

```{r}
consonant_global_stability <- mutate(consonant_stability_class, mglobal = (mmanner + mplace)/2)

ggplot(consonant_global_stability) + 
  geom_bar(aes(x = mglobal, 
               y = reorder(LexifierPhoneme, mglobal),
               fill = class), stat = "identity", show.legend = TRUE) +
  labs(x = "Stability score", y = "Phoneme", fill = "Manner")
```

Calculate the stability of the segments.

```{r}
# qplot(x = duration, y = MeanStability, data = consonant_global_stability, color = duration_group) +
#  geom_smooth(method = "lm") 
```

And we can also increase the number of observations in duration regression by running the analysis at the segment level, rather than on mean stability.

```{r}
# lm(categorical_stability ~ duration + (1|Lexifier) + (1|CreolePhoneme) + (1|Language), data=database)
```



# Word position

Next we ask, does word position influence stability?

First, data prepartion.

```{r}
data_by_position <- database %>% select(Position, LexifierPhoneme, PlaceStability, MannerStability) %>% mutate(Position = tolower(Position))

data_by_position$PlaceStability <- as.numeric(data_by_position$PlaceStability)

data_by_position$MannerStability <- as.numeric(data_by_position$MannerStability)
```

Next, calculate stability for each segment according to its word position.

```{r}
position_stability <- mutate(data_by_position, GlobalStability = (PlaceStability + MannerStability)/2)

position_results <- position_stability %>% group_by(LexifierPhoneme, Position) %>% summarize(m = mean(GlobalStability, na.rm = TRUE))
```

And plot the results for all segments.

```{r}
position_results$Position <- factor(position_results$Position, levels = c('word-initial', 'word-medial', 'word-final'))

ggplot(position_results, aes(x = LexifierPhoneme, y = m, fill = Position)) + 
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

Flip horizontally.

```{r}
ggplot(position_results) + 
  geom_bar(aes(x = m,
               y = reorder(LexifierPhoneme, m),
               fill = Position), 
           stat = "identity", 
           show.legend = TRUE,
           position = "dodge2") +
  labs(x = "Stability score", y = "Phoneme", fill = "Position")
```

Plot the results for segments that show differences.

```{r}
# TODO: update what we show, flip horizontally, change colors
position_results1 <- position_results %>% pivot_wider(names_from = Position, values_from = m)

different_position <- subset(position_results1, position_results1$`word-initial` != position_results1$`word-medial` | position_results1$`word-final` != position_results1$`word-medial`)

different_position_results <- different_position %>% pivot_longer(c(`word-initial`, `word-medial`, `word-final`), names_to = "Position", values_to = "m")

different_position_results$Position <- factor(different_position_results$Position, levels = c('word-initial', 'word-medial', 'word-final'))

ggplot(different_position_results, 
       aes(x = LexifierPhoneme, y = m, fill = Position)) + 
  geom_col(position = position_dodge2(width = 0.9, preserve = "single"))
```

Flip horizontally.

```{r}
# TODO: reorder doesn't work here
ggplot(different_position_results) + 
  geom_bar(aes(x = m,
               y = reorder(LexifierPhoneme, m),
               fill = Position), 
           stat = "identity", 
           show.legend = TRUE,
           position = "dodge2") +
  labs(x = "Stability score", y = "Phoneme", fill = "Position")
```

# References

