---
title: "Supplementary materials for: Consonant stability in Portuguese-based creoles"
author: "Steven Moran and Carlos Silva and Nicholas A. Lester"
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
    toc: true
bibliography: 'bibliography.bib'
---
  
# Overview
  
Supplementary materials for [Consonant Stability in Portuguese-based creoles](https://www.overleaf.com/project/60cdac0dd5871295e0f608fc). Silva, Carlos and Steven Moran. Work in progress.

In this report, we use R [@R] and the following packages [@tidyverse;@knitr;@ggrepel].

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(ggrepel)
```

Load the dataset.

```{r, message=FALSE}
# Because the private token keeps resetting...
# database <- read_csv('https://raw.githubusercontent.com/CreoPhon/CreoPhonPt/main/Creoles.csv?token=AAIGDLS6YB3YSYW7QJWERXDBVYPOE')
# write_csv(database, 'database.csv')
database <- read_csv('database.csv')
```

The data look like this.

```{r}
database %>% head() %>% kable()
```

Let's extend the database with some variables. Duration of contact.

```{r}
database$duration <- database$`EndOfInfluence` - database$`FirstMajorSettlement`
```

Next, a variable of global stability.

```{r}
database <- mutate(database, GlobalStability = (PlaceStability + MannerStability)/2)
# table(database$Language, database$GlobalStability)
```

Categorical variable for duration.

```{r}
database <- database %>% mutate(duration_group = ifelse(duration <= 200, 'short', 'long'))
```

Categorical variable for changes in manner and/or place. Stability in the database is '1' (no change) and '0' change.

```{r}
database <- database %>% mutate(categorical_stability = ifelse(PlaceStability == 1 & MannerStability == 1, 'no manner/no place', NA))

database <- database %>% mutate(categorical_stability = ifelse(PlaceStability == 1 & MannerStability == 0, 'manner/no place', categorical_stability))

database <- database %>% mutate(categorical_stability = ifelse(PlaceStability == 0 & MannerStability == 1, 'no manner/place', categorical_stability))

database <- database %>% mutate(categorical_stability = ifelse(PlaceStability == 0 & MannerStability == 0, 'manner/place', categorical_stability))

table(database$categorical_stability)
```


# Creole stability

Which creoles in the sample are more or less stable overall?

```{r, message=FALSE}
creole_stability <- database %>% group_by(Language, Area, duration, duration_group, ContactConditions) %>% summarize(MeanStability = mean(GlobalStability, na.rm = TRUE))
```

Plot it by area.

```{r}
ggplot(creole_stability) + 
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), fill = Area), 
           stat = "identity", show.legend = TRUE) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score")
```

```{r}
table(creole_stability$Area)
```


Plot it by conditions of contact.

```{r}
ggplot(creole_stability) + 
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), fill = ContactConditions), 
           stat = "identity", show.legend = TRUE) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score")
```

# Duration

We have the overall stability values. What are these in relation to the duration of contact?

There does not seem to be a relationship between overall duration and overall stability.

```{r}
ggplot(creole_stability, aes(x=duration, y=MeanStability)) +
  geom_point()

ggplot(creole_stability, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text(label=creole_stability$Language)

ggplot(creole_stability, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = creole_stability$Language))
```

Results from the simple regression.

```{r}
msd <- lm(MeanStability ~ duration, data=creole_stability)
summary(msd)
```

However, there does seem to be two groups of languages -- ones that belong to "long duration" (>= 400 years) and those that below to "short duration" (<= 200 years).

We can try to split the data and rerun the models, but we note that there are very few data points.

```{r}
tmp_short <- creole_stability %>% filter(duration <= 200)
tmp_long <- creole_stability %>% filter(duration > 200)
```

```{r}
ggplot(tmp_short, aes(x=duration, y=MeanStability)) +
  geom_point()

ggplot(tmp_short, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text(label=tmp_short$Language)

ggplot(tmp_short, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = tmp_short$Language))
```

```{r}
ggplot(tmp_long, aes(x=duration, y=MeanStability)) +
  geom_point()

ggplot(tmp_long, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text(label=tmp_long$Language)

ggplot(tmp_long, aes(x=duration, y=MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = tmp_long$Language))
```


Or perhaps a single model with an interaction term MeanSim ~ duration, group * duration.

```{r}
msd <- lm(MeanStability ~ duration + duration_group * duration, data=creole_stability)
summary(msd)

ggplot(creole_stability, aes(x = duration, y = MeanStability, color = duration_group)) +
  geom_smooth(method = "lm") +
  geom_point()
```

The variability in the two groups is very different. The direction of the effect is interesting: shorter durations yield more stability more consistently. Over time, the variability in mean stability increases. Time is "destabillizing the pattern of stability".

But it looks like you might have something tastier on your hands. The creoles appear to be bouncing back toward the lexifier over time (based on the duration findings; but perhaps I misunderstand).

And we can also increase the number of observations by running the analysis at the segment level, rather than on mean stability.

Exploratory stuff with GAMs.

```{r}
library(mgcv)
# Factorize duration_group
creole_stability$duration_group = as.factor(creole_stability$duration_group)

# Model with an interaction between duration_group and duration
# (with maximum of cubic-spline fit)
msd.gam <- gam(MeanStability ~ duration_group + s(duration, k=3) + s(duration, by=duration_group, k=3), data=creole_stability)

summary(msd.gam)

plot(msd.gam, all.terms=T, shade=T, pages=1)

qqnorm(resid(msd.gam)); qqline(resid(msd.gam))

msd.gam.trimmed <- gam(MeanStability ~ duration_group + s(duration, k=3) + s(duration, by=duration_group, k=3), data=creole_stability %>% filter(MeanStability>0.7))

summary(msd.gam.trimmed)

plot(msd.gam.trimmed, sel=1, shade=T, ylab="Effect on mean stability", xlab="Duration of influence", residuals=T, main="Main effect of duration", cex=5, pch=".", col='dodgerblue'); abline(h=0, lty=2, col="red")

plot(msd.gam.trimmed, sel=2, shade=T, ylab="Effect on mean stability", xlab="Duration of influence", main="Long-term influence", col='dodgerblue'); abline(h=0, lty=2, col="red")

plot(msd.gam.trimmed, sel=3, shade=T, ylab="Effect on mean stability", xlab="Duration of influence", main="Short-term influence", col='dodgerblue'); abline(h=0, lty=2, col="red")

# (dotted lines indicate error)
plot(msd.gam.trimmed, all.terms=T, sel=4, ylab="Effect on mean stability", xlab="Duration group", main="Main effect of duration group")

# checking out the model performance
qqnorm(resid(msd.gam.trimmed)); qqline(resid(msd.gam.trimmed)) # meh
```

Removing the two creoles with the lowest scores produces significant effects. Doesn't seem very reliable, especially given the small sample size. Also, the pattern is strange; a negative trend of duration for long-term influence and a positive one for short-term influence? Also note that the model detected a mean difference between duration groups, with the short group having (slightly) lower mean stability. This appears to be the case but -- again -- so few observations. 

## Duration efects on the segment level

Does duration affect the stability values of specific segments or segment classes?



```{r}
ggplot(database, aes(duration, MannerStability, colour=duration_group)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_wrap(~Class)
```

```{r}
ggplot(database, aes(duration, GlobalStability, colour=duration_group)) + 
  geom_point() + 
  geom_smooth() +
  facet_wrap(~Class)
```






```{r}
ggplot(database, aes(duration, PlaceStability, colour=duration_group)) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  facet_wrap(~LexifierPhoneme)
```

```{r}
ggplot(database, aes(duration, MannerStability, colour=duration_group)) + 
  geom_point() + 
  geom_smooth() +
  facet_wrap(~LexifierPhoneme)
```

```{r}
ggplot(database, aes(duration, GlobalStability, colour=duration_group)) + 
  geom_point() + 
  geom_smooth() +
  facet_wrap(~LexifierPhoneme)
```




# Segment stability

Which segments are the most stable across creoles in the language sample?

We calculate stability of place and manner for each phoneme.

```{r}
place_results <- database %>% group_by(LexifierPhoneme) %>% summarize(mplace = mean(PlaceStability, na.rm = TRUE))
manner_results <- database %>% group_by(LexifierPhoneme) %>% summarize(mmanner = mean(MannerStability, na.rm = TRUE))

consonant_stability <- left_join(place_results, manner_results, by = "LexifierPhoneme")

class <- c("nasal", "rhotic", "lateral", "fricative", "stop", "stop", "fricative", "stop", "stop", "lateral", "nasal", "nasal", "stop", "rhotic", "fricative", "stop", "affricate", "fricative", "fricative")

consonant_stability_class <- cbind(consonant_stability, class)
```

Next, we plot the results.

```{r}
# TODO: fix the class "dots"
ggplot(consonant_stability, aes(y = mmanner, x = mplace, label = LexifierPhoneme, color=class)) +
  geom_point(position= "dodge") + 
  geom_text(aes(label=LexifierPhoneme), hjust=3, vjust=0) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

Here is an alternative view for the global results.

```{r}
consonant_global_stability <- mutate(consonant_stability_class, mglobal = (mmanner + mplace)/2)

ggplot(consonant_global_stability) + 
  geom_bar(aes(x = mglobal, 
               y = reorder(LexifierPhoneme, mglobal),
               fill = class), stat = "identity", show.legend = TRUE) +
  labs(x = "Stability score", y = "Phoneme", fill = "Manner")
```

Calculate the stability of the segments.

```{r}
# qplot(x = duration, y = MeanStability, data = consonant_global_stability, color = duration_group) +
#  geom_smooth(method = "lm") 
```

And we can also increase the number of observations in duration regression by running the analysis at the segment level, rather than on mean stability.

```{r}
library(lmerTest)

# Factorizing
mod.db = database %>%
              as.data.frame() %>%
              mutate(categorical_stability = as.factor(categorical_stability),
              Lexifier = as.factor(Lexifier),
              CreolePhoneme = as.factor(CreolePhoneme),
              Language = as.factor(Language))

# Remove singletons/doubletons
#goodies = names(table(mod.db$CreolePhoneme)>2)

#mod.db = mod.db %>%
#         filter(CreolePhoneme %in% goodies)
```

```{r}
plot(mod.db$categorical_stability, mod.db$duration, notch=T)
```

Hugely skewed in favor of no manner/place (10X as frequent as the next most frequent level; this could cause problems for the models).

```{r}
table(mod.db$categorical_stability)

# Place stability
cat.mod.place = glmer(PlaceStability ~ log(duration) + (1|CreolePhoneme), data=mod.db, family="binomial")

summary(cat.mod.place)

# Manner stability
cat.mod.manner = glmer(MannerStability ~ log(duration) + (1|CreolePhoneme), data=mod.db, family="binomial")

summary(cat.mod.manner)

# Duration group
cat.mod.group = glmer(as.factor(duration_group) ~ PlaceStability + MannerStability + (1|CreolePhoneme), data=mod.db, family="binomial", nAGQ=0)

summary(cat.mod.group)
```

Some indication that place stability is more often associated with shorter periods of influence.

Numerically, the manner/place category has 50% of its observations in the longest durations from the sample. At the same time, no manner/no place is associated with the shortest durations. 


# Word position

Next we ask, does word position influence stability?

First, data preparation.

```{r}
data_by_position <- database %>% select(Position, LexifierPhoneme, PlaceStability, MannerStability) %>% mutate(Position = tolower(Position))

data_by_position$PlaceStability <- as.numeric(data_by_position$PlaceStability)

data_by_position$MannerStability <- as.numeric(data_by_position$MannerStability)
```

Next, calculate stability for each segment according to its word position.

```{r}
position_stability <- mutate(data_by_position, GlobalStability = (PlaceStability + MannerStability)/2)

position_results <- position_stability %>% group_by(LexifierPhoneme, Position) %>% summarize(m = mean(GlobalStability, na.rm = TRUE))
```

And plot the results for all segments.

```{r}
position_results$Position <- factor(position_results$Position, levels = c('word-initial', 'word-medial', 'word-final'))

ggplot(position_results, aes(x = LexifierPhoneme, y = m, fill = Position)) + 
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

Flip horizontally.

```{r}
ggplot(position_results) + 
  geom_bar(aes(x = m,
               y = reorder(LexifierPhoneme, m),
               fill = Position), 
           stat = "identity", 
           show.legend = TRUE,
           position = "dodge2") +
  labs(x = "Stability score", y = "Phoneme", fill = "Position")
```

Plot the results for segments that show differences.

```{r}
# TODO: update what we show, flip horizontally, change colors
position_results1 <- position_results %>% pivot_wider(names_from = Position, values_from = m)

different_position <- subset(position_results1, position_results1$`word-initial` != position_results1$`word-medial` | position_results1$`word-final` != position_results1$`word-medial`)

different_position_results <- different_position %>% pivot_longer(c(`word-initial`, `word-medial`, `word-final`), names_to = "Position", values_to = "m")

different_position_results$Position <- factor(different_position_results$Position, levels = c('word-initial', 'word-medial', 'word-final'))

ggplot(different_position_results, 
       aes(x = LexifierPhoneme, y = m, fill = Position)) + 
  geom_col(position = position_dodge2(width = 0.9, preserve = "single"))
```

Flip horizontally.

```{r}
# TODO: reorder doesn't work here
ggplot(different_position_results) + 
  geom_bar(aes(x = m,
               y = reorder(LexifierPhoneme, m),
               fill = Position), 
           stat = "identity", 
           show.legend = TRUE,
           position = "dodge2") +
  labs(x = "Stability score", y = "Phoneme", fill = "Position")
```


# Conditions of contact

The finding that "slavery has a negative impact on stability" was mainly observational and also literature-based (e.g. Faraclas et al. 2007; Carvalho & Lucchesi 2016; Upper Guinea light creoles = slavery but with lighter contact conditions vs Gulf of Guinea hard creole = slavery and harder contact conditions).

Test whether there's a relation between type of contact situation and overall mean stability.

```{r}
m <- lm(MeanStability ~ ContactConditions, data=creole_stability)
summary(m)

ggplot(creole_stability, aes(x = ContactConditions, y = MeanStability, color = ContactConditions)) +
  geom_smooth(method = "lm") +
  geom_violin()
```

```{r}
m <- lm(MeanStability ~ duration + ContactConditions * duration, data=creole_stability)
summary(m)

ggplot(creole_stability, aes(x = duration, y = MeanStability, color = ContactConditions)) +
  geom_smooth(method = "lm") +
  geom_point()
```


# References

