---
title: "Supplementary materials for: Consonant stability in Portuguese-based creoles"
author: "Steven Moran, Carlos Silva and Nicholas A. Lester" 
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
    toc: true
bibliography: 'bibliography.bib'
---

<!-- FOr the anonymous PDF version, use:

  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
-->


# Overview
  
<!-- Supplementary materials for [Consonant Stability in Portuguese-based creoles](https://www.overleaf.com/project/60cdac0dd5871295e0f608fc). Silva, Carlos and Steven Moran. Work in progress. -->

Supplementary materials for, "Consonant Stability in Portuguese-based creoles". In this report, we provide code in R [@R] and we use these R libraries [@tidyverse;@knitr;@ggrepel;@lmerTest;@mgcv]:

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(ggrepel)
library(lmerTest)
library(mgcv)

# Set the theme for all figures
theme_set(theme_bw())
```

Load the data set.

```{r, message=FALSE}
database <- read_csv("database.csv")
```

The data look like this:

```{r}
database %>%
  head() %>%
  kable()
```

```{r, echo=FALSE, eval=FALSE}
# This is for the PDF version, but it still doesn't display nicely. Skipping it.
database %>%
  head() %>%
knitr::kable(
    format = "latex",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    ) %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```


We extend the database with some additional variables. First, duration of contact.

```{r}
database$duration <- database$`EndOfInfluence` - database$`FirstMajorSettlement`
```

Next, a variable of global stability.

```{r}
database <- mutate(database, GlobalStability = (PlaceStability + MannerStability) / 2)
```

Also, a categorical variable for duration.

```{r}
database <- database %>% 
  mutate(duration_group = ifelse(duration <= 250, "short", "long"))
```

And a categorical variable for changes in manner and/or place. Stability in the database is '1' (no change) and '0' (change).

```{r}
database <- database %>% 
  mutate(categorical_stability = ifelse(PlaceStability == 1 & MannerStability == 1, 
                                        "no manner/no place", NA))

database <- database %>% 
  mutate(categorical_stability = ifelse(PlaceStability == 1 & MannerStability == 0, 
                                        "manner/no place", categorical_stability))

database <- database %>% 
  mutate(categorical_stability = ifelse(PlaceStability == 0 & MannerStability == 1, 
                                        "no manner/place", categorical_stability))

database <- database %>% 
  mutate(categorical_stability = ifelse(PlaceStability == 0 & MannerStability == 0, 
                                        "manner/place", categorical_stability))

table(database$categorical_stability)
```


# Creole stability

Which creoles in the sample are more or less stable overall?

```{r, warning=FALSE, message=FALSE}
creole_stability <- database %>%
  group_by(Language, Area, duration, duration_group, ContactConditions) %>%
  summarize(MeanStability = mean(GlobalStability, na.rm = TRUE))
  
write_csv(creole_stability, 'creole_stability.csv')
```

Plot it by area.

```{r, creole_stability_by_area, warning=FALSE, message=FALSE}
ggplot(creole_stability) +
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), fill = Area),
    stat = "identity", show.legend = TRUE) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score")
```

```{r}
table(creole_stability$Area)
```

Plot it by conditions of contact.

```{r, warning=FALSE, message=FALSE}
ggplot(creole_stability) +
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), 
               fill = ContactConditions), stat = "identity", show.legend = TRUE
  ) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score")
```


# Duration of contact

We have the overall stability values. What are these in relation to the duration of contact?

There does not seem to be a relationship between overall duration and overall stability.

```{r, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = duration, y = MeanStability)) +
  geom_point() +
  xlab("Duration (years)") +
  ylab("Mean stability")
```

```{r, duration_groups_scatter, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = duration, y = MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = creole_stability$Language)) +
  xlab("Duration (years)") +
  ylab("Mean stability")
```

Results from the simple regression.

```{r}
msd <- lm(MeanStability ~ duration, data = creole_stability)
summary(msd)
```

However, there does seem to be two groups of languages -- ones that belong to "long duration" (>= 400 years) and those that below to "short duration" (<= 250 years).

We can try to split the data and rerun the models, but we note that there are very few data points.

```{r}
tmp_short <- creole_stability %>% filter(duration <= 250)
tmp_long <- creole_stability %>% filter(duration > 250)
```

```{r, warning=FALSE, message=FALSE}
ggplot(tmp_short, aes(x = duration, y = MeanStability)) +
  geom_point() +
  xlab("Duration (years)") +
  ylab("Mean stability")

ggplot(tmp_short, aes(x = duration, y = MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = tmp_short$Language)) +
  xlab("Duration (years)") +
  ylab("Mean stability")
```

```{r, warning=FALSE, message=FALSE}
ggplot(tmp_long, aes(x = duration, y = MeanStability)) +
  geom_point() +
  xlab("Duration (years)") +
  ylab("Mean stability")

ggplot(tmp_long, aes(x = duration, y = MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = tmp_long$Language)) +
  xlab("Duration (years)") +
  ylab("Mean stability")
```

A single model with an interaction term MeanSim ~ duration, group * duration.

```{r}
msd <- lm(MeanStability ~ duration + duration_group * duration, data = creole_stability)
summary(msd)
```

```{r, duration_groups, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = duration, y = MeanStability, color = duration_group)) +
  geom_smooth(method = "lm") +
  geom_point() +
  xlab("Duration (years)") +
  ylab("Mean stability") +
  labs(color = "Duration group")
```

The variability in the two groups is very different. The direction of the effect is interesting: shorter durations yield more stability more consistently. Over time, the variability in mean stability increases. Time is "destabillizing the pattern of stability".

But it looks like you might have something tastier on your hands. The creoles appear to be bouncing back toward the lexifier over time (based on the duration findings; but perhaps I misunderstand).

And we can also increase the number of observations by running the analysis at the segment level, rather than on mean stability.

Exploratory analysis with a generalized additive model (GAM).

```{r, warning=FALSE, message=FALSE}
# Factorize duration_group
creole_stability$duration_group <- as.factor(creole_stability$duration_group)

# Model with an interaction between duration_group and duration
# (with maximum of cubic-spline fit)
msd.gam <- gam(MeanStability ~ duration_group + s(duration, k = 3) + 
                 s(duration, by = duration_group, k = 3), data = creole_stability)

summary(msd.gam)

plot(msd.gam, all.terms = T, shade = T, pages = 1)

qqnorm(resid(msd.gam))
qqline(resid(msd.gam))

msd.gam.trimmed <- gam(MeanStability ~ duration_group + s(duration, k = 3) 
                       + s(duration, by = duration_group, k = 3), 
                       data = creole_stability %>% filter(MeanStability > 0.7))

summary(msd.gam.trimmed)

plot(msd.gam.trimmed, sel = 1, shade = T, ylab = "Effect on mean stability", 
     xlab = "Duration of influence", residuals = T, main = "Main effect of duration", 
     cex = 5, pch = ".", col = "dodgerblue")
abline(h = 0, lty = 2, col = "red")

plot(msd.gam.trimmed, sel = 2, shade = T, ylab = "Effect on mean stability", 
     xlab = "Duration of influence", main = "Long-term influence", col = "dodgerblue")
abline(h = 0, lty = 2, col = "red")

plot(msd.gam.trimmed, sel = 3, shade = T, ylab = "Effect on mean stability", 
     xlab = "Duration of influence", main = "Short-term influence", col = "dodgerblue")
abline(h = 0, lty = 2, col = "red")

# (dotted lines indicate error)
plot(msd.gam.trimmed, all.terms = T, sel = 4, ylab = "Effect on mean stability", 
     xlab = "Duration group", main = "Main effect of duration group")

# checking out the model performance
qqnorm(resid(msd.gam.trimmed))
qqline(resid(msd.gam.trimmed)) # meh
```

Removing the two creoles with the lowest scores produces significant effects. This doesn't seem very reliable though, especially given the small sample size. Also, the pattern is strange: a negative trend of duration for long-term influence and a positive one for short-term influence? Note that the model detected a mean difference between duration groups, with the short group having (slightly) lower mean stability. This appears to be the case -- but again -- we have so few observations. 


## Duration effects on the segment level

Does duration affect the stability values of specific segments or segment classes?

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, MannerStability, colour = duration_group)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~Class) +
  xlab("Duration (years)") +
  ylab("Manner stability") +
  labs(color = "Duration")
```

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, GlobalStability, colour = duration_group)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~Class) +
  xlab("Duration (years)") +
  ylab("Global stability") +
  labs(color = "Duration")
```

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, PlaceStability, colour = duration_group)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~LexifierPhoneme) +
  xlab("Duration (years)") +
  ylab("Place stability") +
  labs(color = "Duration")
```

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, MannerStability, colour = duration_group)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~LexifierPhoneme) +
  xlab("Duration (years)") +
  ylab("Mean stability") +
  labs(color = "Duration")
```

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, GlobalStability, colour = duration_group)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~LexifierPhoneme) +
  xlab("Duration (years)") +
  ylab("Global stability") +
  labs(color = "Duration")
```


# Segment stability

Which segments are the most stable across creoles in the language sample?

We calculate stability of place and manner for each phoneme.

```{r, warning=FALSE, message=FALSE}
place_results <- database %>%
  group_by(LexifierPhoneme) %>%
  summarize(mplace = mean(PlaceStability, na.rm = TRUE))
manner_results <- database %>%
  group_by(LexifierPhoneme) %>%
  summarize(mmanner = mean(MannerStability, na.rm = TRUE))

consonant_stability <- left_join(place_results, manner_results, by = "LexifierPhoneme")

class <- c("stop", "stop", "fricative", "stop", "stop", "lateral","nasal", "nasal", "stop", "rhotic","fricative", "stop","affricate", "fricative", "fricative","nasal", "rhotic", "lateral", "fricative")

consonant_stability_class <- cbind(consonant_stability, class)
```

Next, we plot the results.

```{r, stability_by_consonant, warning=FALSE, message=FALSE}
ggplot(consonant_stability, aes(y = mmanner, x = mplace)) +
  geom_point(position = "dodge", aes(color = class)) +
  geom_text_repel(aes(label = LexifierPhoneme), size = 4) +
  xlab("Mean stability (place of articulation)") +
  ylab("Mean stability (manner of articulation)") +
  labs(color = "Segment class")
```

A linear model to assess the relationship between manner and place stability


```{r, warning=FALSE, message=FALSE}
lm_manner_place <- lm(mplace~mmanner, data=consonant_stability)

summary(lm_manner_place)

manner_place_lmplot <- ggplot(consonant_stability, aes(y = mmanner, x = mplace, label = LexifierPhoneme)) +
  geom_point(position= "dodge") + 
  geom_smooth(method = lm)+
  geom_text(aes(label=LexifierPhoneme), hjust=3, vjust=0)

print(manner_place_lmplot + labs(y = "Manner Stability", x = "Place Stability"))
```

Here is an alternative view for the global results.

```{r, warning=FALSE, message=FALSE}
consonant_global_stability <- mutate(consonant_stability_class, 
                                     mglobal = (mmanner + mplace) / 2)

write.csv(consonant_global_stability, "consonant_global_stability.csv", row.names=FALSE)

ggplot(consonant_global_stability) +
  geom_bar(aes(
    x = mglobal,
    y = reorder(LexifierPhoneme, mglobal),
    fill = class
  ), stat = "identity", show.legend = TRUE) +
  labs(x = "Stability score", y = "Phoneme", fill = "Manner")
```

Check for class effects on the global stability of consonants

```{r, warning=FALSE, message=FALSE}

ggplot(consonant_global_stability, aes(x = class, y = mglobal, fill = class)) +
  geom_smooth(method = "lm") +
  geom_violin() +
   geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5)
```

Now, just plotting the relation manner to manner
```{r, warning=FALSE, message=FALSE}
ggplot(consonant_global_stability, aes(x = class, y = mmanner, fill = class)) +
  geom_smooth(method = "lm") +
  geom_violin() +
   geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5) +
  theme(legend.position="none") +
 stat_summary(fun.y=mean, geom="point", size=2, shape=16)
```



Check place effects on the global stability of the consonants

```{r, warning=FALSE, message=FALSE}

place <- c("labial", "alveolar", "labiodental", "velar", "velar", "alveolar", "labial", "alveolar", "labial", "alveolar", "alveolar", "alveolar", "palatal", "labiodental", "alveolar", "palatal", "alveolar", "palatal", "palatal")

consonant_stability_place <- cbind(consonant_global_stability, place)

ggplot(consonant_stability_place, aes(x = place, y = mglobal, fill = place)) +
  geom_smooth(method = "lm") +
  geom_violin() +
   geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5) +
  theme(legend.position="none")

```
Now, just with the mean for Place stability

```{r, warning=FALSE, message=FALSE}
ggplot(consonant_stability_place, aes(x = place, y = mplace, fill = place)) +
  geom_smooth(method = "lm") +
  geom_violin()  +
   geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5)  +
  theme(legend.position="none") +
  ylab("Place stability score") +
  xlab("Place of articulation") +
 stat_summary(fun.y=mean, geom="point", size=2, shape=16)
```

Calculate the stability of the segments.

```{r}
# qplot(x = duration, y = MeanStability, data = consonant_global_stability, color = duration_group) +
#  geom_smooth(method = "lm") 
```
And we can also increase the number of observations in duration regression by running the analysis at the segment level, rather than on mean stability.

```{r}
# Factorizing
mod.db <- database %>%
  as.data.frame() %>%
  mutate(
    categorical_stability = as.factor(categorical_stability),
    Lexifier = as.factor(Lexifier),
    CreolePhoneme = as.factor(CreolePhoneme),
    Language = as.factor(Language)
  )

# Remove singletons/doubletons
# goodies = names(table(mod.db$CreolePhoneme)>2)

# mod.db = mod.db %>%
#         filter(CreolePhoneme %in% goodies)
```

```{r, warning=FALSE, message=FALSE}
plot(mod.db$categorical_stability, mod.db$duration, notch = T)
```

Hugely skewed in favor of no manner/place (10x as frequent as the next most frequent level; this could cause problems for the models).

```{r}
table(mod.db$categorical_stability)

# Place stability
cat.mod.place <- glmer(PlaceStability ~ log(duration) + (1 | CreolePhoneme), 
                       data = mod.db, family = "binomial")

summary(cat.mod.place)

# Manner stability
cat.mod.manner <- glmer(MannerStability ~ log(duration) + (1 | CreolePhoneme), 
                        data = mod.db, family = "binomial")

summary(cat.mod.manner)

# Duration group
cat.mod.group <- glmer(as.factor(duration_group) ~ PlaceStability + 
                         MannerStability + (1 | CreolePhoneme), 
                       data = mod.db, family = "binomial", nAGQ = 0)

summary(cat.mod.group)
```

Some indication that place stability is more often associated with shorter periods of influence.

Numerically, the manner/place category has 50% of its observations in the longest duration from the sample. At the same time, no manner/no place is associated with the shortest duration. 


# Word position

Next we ask, does word position influence stability?

First, data preparation.

```{r, warning=FALSE, message=FALSE}
data_by_position <- database %>%
  select(Position, LexifierPhoneme, PlaceStability, MannerStability) %>%
  mutate(Position = tolower(Position))

data_by_position$PlaceStability <- as.numeric(data_by_position$PlaceStability)

data_by_position$MannerStability <- as.numeric(data_by_position$MannerStability)
```

Next, calculate stability for each segment according to its word position.

```{r, warning=FALSE, message=FALSE}
position_stability <- mutate(data_by_position, GlobalStability = 
                               (PlaceStability + MannerStability) / 2)

position_results <- position_stability %>%
  group_by(LexifierPhoneme, Position) %>%
  summarize(m = mean(GlobalStability, na.rm = TRUE))
```

And plot the results for all segments.

```{r, warning=FALSE, message=FALSE}
position_results$Position <- factor(position_results$Position, 
                                    levels = c("word-initial", 
                                               "word-medial", 
                                               "word-final"))

ggplot(position_results, aes(x = LexifierPhoneme, y = m, fill = Position)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  labs(x = "Lexifier phoneme", y = "Mean stability", fill = "Word position")
```

Flip horizontally.

```{r, correlation_stability_word_position, warning=FALSE, message=FALSE}
ggplot(position_results) +
  geom_bar(
    aes(
      x = m,
      y = reorder(LexifierPhoneme, m),
      fill = Position
    ),
    stat = "identity",
    show.legend = TRUE,
    position = "dodge2"
  ) +
  labs(x = "Stability score", y = "Phoneme", fill = "Word position")
```

Plot the results for segments that show differences.

```{r, warning=FALSE, message=FALSE}
position_results1 <- position_results %>% 
  pivot_wider(names_from = Position, values_from = m)

different_position <- subset(position_results1, position_results1$`word-initial` 
                             != position_results1$`word-medial` | 
                               position_results1$`word-final` 
                             != position_results1$`word-medial`)

different_position_results <- different_position %>% 
  pivot_longer(c(`word-initial`, `word-medial`, `word-final`), 
               names_to = "Position", values_to = "m")

different_position_results$Position <- factor(different_position_results$Position, 
                                              levels = c("word-initial", 
                                                         "word-medial", 
                                                         "word-final"))

ggplot(
  different_position_results,
  aes(x = LexifierPhoneme, y = m, fill = Position)
) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  labs(x = "Lexifier phoneme", y = "Mean stability", fill = "Word position")
```

Flip horizontally.

```{r, warning=FALSE, message=FALSE}
ggplot(different_position_results) +
  geom_bar(
    aes(
      x = m,
      y = reorder(LexifierPhoneme, m),
      fill = Position
    ),
    stat = "identity",
    show.legend = TRUE,
    position = "dodge2"
  ) +
  labs(x = "Stability score", y = "Phoneme", fill = "Word position")
```


# Conditions of contact

The finding that "slavery has a negative impact on stability" was mainly observational and also literature-based (e.g. @Faraclasetal07; @CarvalhoLucchesi16; Upper Guinea light creoles = slavery but with lighter contact conditions versus Gulf of Guinea hard creole = slavery and harder contact conditions).

Test whether there's a relation between type of contact situation and overall mean stability.

Linear model
```{r}
m <- lm(MeanStability ~ ContactConditions, data = creole_stability)
summary(m)
```

```{r, conditions_violin, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = ContactConditions, y = MeanStability, 
                             fill = ContactConditions)) +
  geom_smooth(method = "lm") +
  geom_violin() +
  geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5)  +
  theme(legend.position="none")
 ```

```{r, warning=FALSE, message=FALSE}
m <- lm(MeanStability ~ duration + ContactConditions * duration, 
        data = creole_stability)
summary(m)
```

```{r, duration_groups_geom, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = duration, y = MeanStability, 
                             color = ContactConditions)) +
  geom_smooth(method = "lm") +
  geom_point() +
  xlab("Duration (years)") +
  ylab("Mean stability") +
  labs(color = "Contact conditions")
```


# References
