---
title: "Supplementary materials for: Consonant stability in Portuguese-based creoles"
author: "Steven Moran and Carlos Silva" 
date: "(`r format(Sys.time(), '%d %B, %Y')`)"
output:
  github_document:
    toc: true
    number_sections: true
bibliography: 'bibliography.bib'
---

<!-- For the anonymous PDF version, use:

  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
-->


# Overview
  
<!-- Supplementary materials for [Consonant Stability in Portuguese-based creoles](https://www.overleaf.com/project/60cdac0dd5871295e0f608fc). Silva, Carlos and Steven Moran. Work in progress. -->

Supplementary materials for, "Consonant Stability in Portuguese-based creoles", by Carlos Silva and Steven Moran, to appear in [Diachronica](https://benjamins.com/catalog/dia).

These supplementary materials are available online at:

* https://github.com/CreoPhon/consonant_stability_portuguese_creoles

In this report, we provide code in R [@R] and we use these R libraries [@tidyverse;@knitr;@ggrepel;@lmerTest;@mgcv;@PMCMRplus;@ggpubr;@rstatix;@prabclus;@maps]:

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(ggrepel)
library(lmerTest)
library(mgcv)
library(PMCMRplus)
library(ggpubr)
library(rstatix)
library(stats)
library(prabclus)
library(maps)

# Set the theme for all figures
theme_set(theme_bw())
```

Load the data set.

```{r, message=FALSE}
database <- read_csv("database.csv")
```

<!-- Doesn't appear nicely in PDF
The data look like this:

```{r}
database %>%
  head() %>%
  kable()
```

-->

```{r, echo=FALSE, eval=FALSE}
# This is for the PDF version, but it still doesn't display nicely. Skipping it.
database %>%
  head() %>%
knitr::kable(
    format = "latex",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
    ) %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```


We extend the database with some additional variables. First, duration of contact.

```{r}
database$duration <- database$`EndOfInfluence` - database$`FirstMajorSettlement`
```

Next, a variable of global stability.

```{r}
database <- mutate(database, GlobalStability = (PlaceStability + MannerStability) / 2)
```

Also, a categorical variable for duration.

```{r}
database <- database %>% 
  mutate(duration_group = ifelse(duration <= 250, "short", "long"))
```

And a categorical variable for changes in manner and/or place. Stability in the database is '1' (no change) and '0' (change).

```{r}
database <- database %>% 
  mutate(categorical_stability = ifelse(PlaceStability == 1 & MannerStability == 1, 
                                        "no manner/no place", NA))

database <- database %>% 
  mutate(categorical_stability = ifelse(PlaceStability == 1 & MannerStability == 0, 
                                        "manner/no place", categorical_stability))

database <- database %>% 
  mutate(categorical_stability = ifelse(PlaceStability == 0 & MannerStability == 1, 
                                        "no manner/place", categorical_stability))

database <- database %>% 
  mutate(categorical_stability = ifelse(PlaceStability == 0 & MannerStability == 0, 
                                        "manner/place", categorical_stability))

table(database$categorical_stability)
```

# Language sample

Here are the languages plotted on a world map.

```{r, message=FALSE}
locations <- read_csv('creole_locations.csv')

ggplot(data = locations, aes(x = Longitude, y = Latitude)) + 
  borders("world", colour="gray50", fill="gray50") + 
  geom_point()
```


# Creole stability

Which creoles in the sample are more or less stable overall?

```{r, warning=FALSE, message=FALSE}
creole_stability <- database %>%
  group_by(Language, Area, duration, duration_group, ContactConditions) %>%
  summarize(MeanStability = mean(GlobalStability, na.rm = TRUE))
  
write_csv(creole_stability, 'creole_stability.csv')
```

Plot it by area.

```{r, creole_stability_by_area, warning=FALSE, message=FALSE}
ggplot(creole_stability) +
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), 
               fill = Area), stat = "identity", show.legend = TRUE)  +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score")
```

Plot it by area with a color-blind friendly palette.

```{r}
# cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
cbPalette <- c("#56B4E9", "#F0E442", "#999999", "#E69F00", "#009E73", "#0072B2", "#D55E00", "#CC79A7")
# cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r, creole_stability_by_area_cbf}
ggplot(creole_stability) +
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), 
               fill = Area), stat = "identity", show.legend = TRUE)  +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score") + 
  scale_fill_manual(values=cbPalette)
```

```{r, echo=FALSE, eval=FALSE}
ggplot(creole_stability) +
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), 
               fill = Area), stat = "identity", show.legend = TRUE)  +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score") + 
  scale_fill_manual(values=cbbPalette)
```

```{r}
table(creole_stability$Area)
```


## Conditions of contact

We have the overall stability values. What are these in relation to the conditions of contact?

The finding that "slavery has a negative impact on stability" was mainly observational (e.g. @Faraclasetal07; @CarvalhoLucchesi16). Upper Guinea are considered "light" creoles, i.e., enslaved people with "lighter" contact conditions. Gulf of Guinea are considered "hard" creole, i.e., enslaved people under "harder" contact conditions.


```{r, warning=FALSE, message=FALSE}
ggplot(creole_stability) +
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), 
               fill = ContactConditions), stat = "identity", show.legend = TRUE
  ) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score", fill = "Conditions of contact")
```

```{r, creole_stability_cbf}
ggplot(creole_stability) +
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), 
               fill = ContactConditions), stat = "identity", show.legend = TRUE
  ) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score", fill = "Conditions of contact") + 
  scale_fill_manual(values=cbPalette)
```

```{r, echo=FALSE, eval=FALSE}
ggplot(creole_stability) +
  geom_bar(aes(x = MeanStability, y = reorder(Language, MeanStability), 
               fill = ContactConditions), stat = "identity", show.legend = TRUE
  ) +
  theme(axis.title.y = element_blank()) +
  labs(x = "Stability score", fill = "Conditions of contact") + 
  scale_fill_manual(values=cbbPalette)
```

Test whether there is a relation between type of contact situation and overall mean stability with a linear model.

```{r}
m <- lm(MeanStability ~ ContactConditions, data = creole_stability)
summary(m)
```

Visualize the results.

```{r, conditions_violin, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = ContactConditions, y = MeanStability, 
                             fill = ContactConditions)) +
  geom_smooth(method = "lm", colour="black") +
  geom_violin() +
  geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5)  +
  theme(legend.position="none") +
  labs(y = "Stability score", x = "Conditions of contact")
```

```{r, conditions_violin_cbf}
ggplot(creole_stability, aes(x = ContactConditions, y = MeanStability, 
                             fill = ContactConditions)) +
  geom_smooth(method = "lm", colour="black") +
  geom_violin() +
  geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5)  +
  theme(legend.position="none") +
  labs(y = "Stability score", x = "Conditions of contact") +
  scale_fill_manual(values=cbPalette)
```

Statistical summary.

```{r, warning=FALSE, message=FALSE}
m <- lm(MeanStability ~ duration + ContactConditions * duration, 
        data = creole_stability)
summary(m)
```

```{r, duration_groups_geom, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = duration, y = MeanStability, color = ContactConditions)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  xlab("Duration (years)") +
  ylab("Mean stability") +
  labs(color = "Contact conditions")
```

```{r, duration_groups_geom_cbf}
ggplot(creole_stability, aes(x = duration, y = MeanStability, fill = ContactConditions)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  xlab("Duration (years)") +
  ylab("Mean stability") +
  labs(fill = "Contact conditions") +
  scale_fill_manual(values=cbPalette)
```


## Duration of contact

We have the overall stability values. What are these in relation to the duration of contact?

There does not seem to be a relationship between overall duration and overall stability.

```{r, duration_groups_scatter, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = duration, y = MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = creole_stability$Language)) +
  xlab("Duration of contact (years)") +
  ylab("Stability score")
```

Results from a linear regression.

```{r}
msd <- lm(MeanStability ~ duration, data = creole_stability)
summary(msd)
```

However, there does seem to be two groups of languages -- ones that belong to "long duration" (>= 400 years) and those that belong to "short duration" (<= 250 years).

We split the data and rerun the models, but we note that there are very few data points.

```{r}
tmp_short <- creole_stability %>% filter(duration <= 250)
tmp_long <- creole_stability %>% filter(duration > 250)
```

```{r, warning=FALSE, message=FALSE}
ggplot(tmp_short, aes(x = duration, y = MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = tmp_short$Language)) +
  xlab("Duration (years)") +
  ylab("Mean stability")
```

```{r, warning=FALSE, message=FALSE}
ggplot(tmp_long, aes(x = duration, y = MeanStability)) +
  geom_point() +
  geom_text_repel(aes(label = tmp_long$Language)) +
  xlab("Duration (years)") +
  ylab("Mean stability")
```

Here is a single model with an interaction term MeanSim ~ duration, group * duration.

```{r}
msd <- lm(MeanStability ~ duration + duration_group * duration, data = creole_stability)
summary(msd)
```

```{r, duration_groups, warning=FALSE, message=FALSE}
ggplot(creole_stability, aes(x = duration, y = MeanStability, color = duration_group)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  xlab("Duration of contact (years)") +
  ylab("Stability score") +
  labs(color = "Group")
```

```{r, duration_groups_cbf}
ggplot(creole_stability, aes(x = duration, y = MeanStability, fill = duration_group)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  xlab("Duration of contact (years)") +
  ylab("Stability score") +
  labs(color = "Group") +
  scale_fill_manual(values=cbPalette)
```


The variability in the two groups is very different. The direction of the effect is interesting: shorter durations yield more stability more consistently. Over time, the variability in mean stability increases. Time is "destabillizing the pattern of stability" (pc Nicholas Lester). 

And we can also increase the number of observations by running the analysis at the segment level, rather than on mean stability.

Here is an exploratory analysis with a generalized additive model (GAM).

```{r, warning=FALSE, message=FALSE}
# Factorize duration_group
creole_stability$duration_group <- as.factor(creole_stability$duration_group)

# Model with an interaction between duration_group and duration
# (with maximum of cubic-spline fit)
msd.gam <- gam(MeanStability ~ duration_group + s(duration, k = 3) + 
                 s(duration, by = duration_group, k = 3), data = creole_stability)

summary(msd.gam)

plot(msd.gam, all.terms = T, shade = T, pages = 1)

qqnorm(resid(msd.gam))
qqline(resid(msd.gam))

msd.gam.trimmed <- gam(MeanStability ~ duration_group + s(duration, k = 3) 
                       + s(duration, by = duration_group, k = 3), 
                       data = creole_stability %>% filter(MeanStability > 0.7))

summary(msd.gam.trimmed)

plot(msd.gam.trimmed, sel = 1, shade = T, ylab = "Effect on mean stability", 
     xlab = "Duration of influence", residuals = T, main = "Main effect of duration", 
     cex = 5, pch = ".", col = "dodgerblue")
abline(h = 0, lty = 2, col = "red")

plot(msd.gam.trimmed, sel = 2, shade = T, ylab = "Effect on mean stability", 
     xlab = "Duration of influence", main = "Long-term influence", col = "dodgerblue")
abline(h = 0, lty = 2, col = "red")

plot(msd.gam.trimmed, sel = 3, shade = T, ylab = "Effect on mean stability", 
     xlab = "Duration of influence", main = "Short-term influence", col = "dodgerblue")
abline(h = 0, lty = 2, col = "red")

# (dotted lines indicate error)
plot(msd.gam.trimmed, all.terms = T, sel = 4, ylab = "Effect on mean stability", 
     xlab = "Duration group", main = "Main effect of duration group")

# checking out the model performance
qqnorm(resid(msd.gam.trimmed))
qqline(resid(msd.gam.trimmed)) # meh
```

Removing the two creoles with the lowest scores produces significant effects. This does not seem very reliable though, especially given the small sample size. Also, the pattern is strange: a negative trend of duration for long-term influence and a positive one for short-term influence? Note that the model detected a mean difference between duration groups, with the short group having (slightly) lower mean stability. This appears to be the case -- but again -- there are very few observations. 


## Duration effects on the segment level

Does duration affect the stability values of specific segments or segment classes?

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, MannerStability, colour = duration_group, shape = duration_group)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~Class) +
  xlab("Duration (years)") +
  ylab("Manner stability") +
  labs(color = "Duration")
```

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, GlobalStability, colour = duration_group, shape = duration_group)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~Class) +
  xlab("Duration (years)") +
  ylab("Global stability") +
  labs(color = "Duration")
```

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, PlaceStability, colour = duration_group, shape = duration_group)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~LexifierPhoneme) +
  xlab("Duration (years)") +
  ylab("Place stability") +
  labs(color = "Duration")
```

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, MannerStability, colour = duration_group, shape = duration_group)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~LexifierPhoneme) +
  xlab("Duration (years)") +
  ylab("Mean stability") +
  labs(color = "Duration")
```

```{r, warning=FALSE, message=FALSE}
ggplot(database, aes(duration, GlobalStability, colour = duration_group, shape = duration_group)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(~LexifierPhoneme) +
  xlab("Duration (years)") +
  ylab("Global stability") +
  labs(color = "Duration")
```


## Jaccard distance between inventories

Here we explore the Jaccard distance between phonological inventories.

```{r, warning=FALSE, message=FALSE}
df_jac <- read_csv("Inventories.csv")
#df_jac <- df_jac %>% subset(Category != 'creole') 
df_jac <- df_jac %>% dplyr::select(c('Language','Phoneme'))
df_jac$presence <- 1
df_wide <- df_jac %>% spread(Phoneme, presence)
df_wide <- df_wide %>% replace(is.na(.), 0)
head(df_wide)


tmp <- df_wide %>%
  column_to_rownames(var = "Language")
jac_dist <- jaccard(t(tmp))

```
The Jaccard distance values were manually extracted into a new table, so we could visualize those values according to the relevant language in contact (see file `jaccard_results.csv`). Then, we created a new table which summaries those results for creoles and joins their stability values, so we can assess if there is or if there is not a correlation between the Jaccard distances and the overall stability os creoles.

```{r}
df_cor <- read.csv("jaccard_summary.csv")
```

We create a linear model of inventory distance creoles ~ Portuguese.

```{r, warning=FALSE, message=FALSE}

cl <- lm(stability ~ lex_crio, data=df_cor)
summary(cl)

ggplot(df_cor, aes(x = stability, y = lex_crio, label = Language)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  geom_text_repel(aes(label = Language)) +
  xlab("Stability") +
  ylab("Lex crio") # CS: FIX THIS
```

Next, inventory distance substrates ~ Portuguese.

```{r, warning=FALSE, message=FALSE}
sl_mean <- lm(stability ~ lex_subs_mean, data=df_cor)
summary(sl_mean)

ggplot(df_cor, aes(x = stability, y = lex_subs_mean, label = Language)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  geom_text_repel(aes(label = Language)) +
  xlab("Stability") +
  ylab("Lex subs mean") # CS: FIX THIS

```

Next, inventory distance substrates ~ creoles.


```{r, warning=FALSE, message=FALSE}
sc_mean <- lm(stability ~ crio_subs_mean, data=df_cor)
summary(sc_mean)

ggplot(df_cor, aes(x = stability, y = crio_subs_mean, label = Language)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  geom_text_repel(aes(label = Language)) +
  xlab("Stability") +
  ylab("Crio subs mean") # CS: FIX THIS

```


# Consonant stability

Which segments are the most stable across creoles in the language sample? We calculate stability of place and manner for each phoneme.

```{r, warning=FALSE, message=FALSE}
place_results <- database %>%
  group_by(LexifierPhoneme) %>%
  summarize(mplace = mean(PlaceStability, na.rm = TRUE))
manner_results <- database %>%
  group_by(LexifierPhoneme) %>%
  summarize(mmanner = mean(MannerStability, na.rm = TRUE))

consonant_stability <- left_join(place_results, manner_results, by = "LexifierPhoneme")

class <- c("stop", "stop", "fricative", "stop", "stop", "lateral",
           "nasal", "nasal", "stop", "rhotic","fricative", "stop",
           "affricate", "fricative", "fricative","nasal", "rhotic", 
           "lateral", "fricative")

consonant_stability_class <- cbind(consonant_stability, class)
```

Next, we plot the results.

```{r, stability_by_consonant, warning=FALSE, message=FALSE}
ggplot(consonant_stability, aes(y = mmanner, x = mplace)) +
  geom_point(position = "dodge", aes(color = class)) +
  geom_text_repel(aes(label = LexifierPhoneme), size = 4) +
  xlab("Mean stability (place of articulation)") +
  ylab("Mean stability (manner of articulation)") +
  labs(color = "Segment class")
```

We make a linear model to assess the relationship between manner and place stability.

```{r, manner_place_cor, warning=FALSE, message=FALSE}
lm_manner_place <- lm(mplace~mmanner, data=consonant_stability)

summary(lm_manner_place)

manner_place_lmplot <- ggplot(consonant_stability, 
                              aes(y = mmanner, x = mplace, 
                                  label = LexifierPhoneme)) +
  geom_point(position= "dodge") + 
  geom_smooth(method = lm) #+
  #geom_text(aes(label=LexifierPhoneme), hjust=3, vjust=0)

print(manner_place_lmplot + labs(y = "Manner stability", x = "Place stability")) + 
  geom_text_repel(aes(label=LexifierPhoneme), hjust=3, vjust=1)
```

Here is an alternative view for the global results.

```{r, consonant_stability, warning=FALSE, message=FALSE}
consonant_global_stability <- mutate(consonant_stability_class, 
                                     mglobal = (mmanner + mplace) / 2)

write.csv(consonant_global_stability, "consonant_global_stability.csv", row.names=FALSE)

ggplot(consonant_global_stability) +
  geom_bar(aes(
    x = mglobal,
    y = reorder(LexifierPhoneme, mglobal),
    fill = class
  ), stat = "identity", show.legend = TRUE) +
  labs(x = "Stability score", y="", fill = "Class")
```

```{r, consonant_stability_cbf}
ggplot(consonant_global_stability) +
  geom_bar(aes(
    x = mglobal,
    y = reorder(LexifierPhoneme, mglobal),
    fill = class
  ), stat = "identity", show.legend = TRUE) +
  labs(x = "Stability score", y="", fill = "Class") +
  scale_fill_manual(values=cbPalette)
```


## Manner stability

Check for class effects on the global stability of consonants.

```{r, warning=FALSE, message=FALSE}

ggplot(consonant_global_stability, aes(x = class, y = mglobal, fill = class)) +
  geom_smooth(method = "lm", colour="black") +
  geom_violin() +
   geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5) +
  xlab("Segment calss") +
  ylab("Mean global score") # CS: FIX THIS

```

Now, just plotting the relation manner to manner.

```{r, manner_manner, warning=FALSE, message=FALSE}
ggplot(consonant_global_stability, aes(x = class, y = mmanner, fill = class)) +
  geom_smooth(method = "lm", colour="black") +
  geom_violin() +
   geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5) +
  theme(legend.position="none") +
  ylab("Manner stability score") +
  xlab("Manner of articulation") +
 stat_summary(fun.y=mean, geom="point", size=2, shape=16)
```

## Place stability

We check place effects on the global stability of the consonants.

```{r, warning=FALSE, message=FALSE}
place <- c("labial", "alveolar", "labiodental", "velar", "velar", 
           "alveolar", "labial", "alveolar", "labial", "alveolar", 
           "alveolar", "alveolar", "palatal", "labiodental", "alveolar", 
           "palatal", "alveolar", "palatal", "palatal")

consonant_stability_place <- cbind(consonant_global_stability, place)

ggplot(consonant_stability_place, aes(x = place, y = mglobal, fill = place)) +
  geom_smooth(method = "lm", colour="black") +
  geom_violin() +
   geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5) +
  theme(legend.position="none") +
  xlab("Place of articulation") +
  ylab("Mean global score") # CS: FIX THIS


```
Now, just with the mean for place stability.

```{r, place_place, warning=FALSE, message=FALSE}
ggplot(consonant_stability_place, aes(x = place, y = mplace, fill = place)) +
  geom_smooth(method = "lm", colour="black") +
  geom_violin()  +
   geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5)  +
  theme(legend.position="none") +
  ylab("Place stability score") +
  xlab("Place of articulation") +
 stat_summary(fun.y=mean, geom="point", size=2, shape=16)
```
We can also increase the number of observations in duration regression by running the analysis at the segment level, rather than on mean stability.

```{r}
# Factorizing
mod.db <- database %>%
  as.data.frame() %>%
  mutate(
    categorical_stability = as.factor(categorical_stability),
    Lexifier = as.factor(Lexifier),
    CreolePhoneme = as.factor(CreolePhoneme),
    Language = as.factor(Language)
  )

# Remove singletons/doubletons
# goodies = names(table(mod.db$CreolePhoneme)>2)

# mod.db = mod.db %>%
#         filter(CreolePhoneme %in% goodies)
```

```{r, warning=FALSE, message=FALSE}
plot(mod.db$categorical_stability, mod.db$duration, notch = T)
```

Hugely skewed in favor of no manner/place (10x as frequent as the next most frequent level; note this may cause problems for the models).

```{r}
table(mod.db$categorical_stability)

# Place stability
cat.mod.place <- glmer(PlaceStability ~ log(duration) + (1 | CreolePhoneme), 
                       data = mod.db, family = "binomial")

summary(cat.mod.place)

# Manner stability
cat.mod.manner <- glmer(MannerStability ~ log(duration) + (1 | CreolePhoneme), 
                        data = mod.db, family = "binomial")

summary(cat.mod.manner)

# Duration group
cat.mod.group <- glmer(as.factor(duration_group) ~ PlaceStability + 
                         MannerStability + (1 | CreolePhoneme), 
                       data = mod.db, family = "binomial", nAGQ = 0)

summary(cat.mod.group)
```

There is some indication that place stability is more often associated with shorter periods of influence.

Numerically, the manner/place category has 50% of its observations in the longest duration from the sample. At the same time, no manner/no place is associated with the shortest duration. 


## Word position

Next we ask, does word position influence stability?

First, data preparation.

```{r, warning=FALSE, message=FALSE}
data_by_position <- database %>%
  dplyr::select(Position, LexifierPhoneme, PlaceStability, MannerStability) %>%
  mutate(Position = tolower(Position))

data_by_position$PlaceStability <- as.numeric(data_by_position$PlaceStability)

data_by_position$MannerStability <- as.numeric(data_by_position$MannerStability)
```

Next, calculate stability for each segment according to its word position.

```{r, warning=FALSE, message=FALSE}
position_stability <- mutate(data_by_position, GlobalStability = 
                               (PlaceStability + MannerStability) / 2)

position_results <- position_stability %>%
  group_by(LexifierPhoneme, Position) %>%
  summarize(m = mean(GlobalStability, na.rm = TRUE))
```

And plot the results for all segments.

```{r, position_stability_cbf, warning=FALSE, message=FALSE}
position_results$Position <- factor(position_results$Position, 
                                    levels = c("word-initial", 
                                               "word-medial", 
                                               "word-final"),
                                    ordered = TRUE)

ggplot(position_results, aes(x = reorder(LexifierPhoneme, -m), y = m, fill = Position)) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  ) +
  labs(x = "Lexifier phoneme", y = "Mean stability", fill = "Word position")
```

Flip horizontally.

```{r, position_stability, warning=FALSE, message=FALSE}
ggplot(position_results) +
  geom_bar(
    aes(
      x = m,
      y = reorder(LexifierPhoneme, -m),
      # y = LexifierPhoneme,
      fill = Position
    ),
    stat = "identity",
    show.legend = TRUE,
    position = "dodge2"
  ) +
  labs(x = "Stability score", y = "", fill = "Position")
```

```{r, position_stability_gray, warning=FALSE, message=FALSE}
ggplot(position_results) +
  geom_bar(
    aes(
      x = m,
      y = reorder(LexifierPhoneme, m),
      fill = Position
    ),
    stat = "identity",
    show.legend = TRUE,
    position = "dodge2"
  ) +
  labs(x = "Stability score", y = "", fill = "Position") +
  scale_fill_grey(start = 0.8, end = 0.2) 
```

Plot the results for segments that show differences.

```{r, warning=FALSE, message=FALSE}
position_results1 <- position_results %>% 
  pivot_wider(names_from = Position, values_from = m)

different_position <- subset(position_results1, position_results1$`word-initial` 
                             != position_results1$`word-medial` | 
                               position_results1$`word-final` 
                             != position_results1$`word-medial`)

different_position_results <- different_position %>% 
  pivot_longer(c(`word-initial`, `word-medial`, `word-final`), 
               names_to = "Position", values_to = "m")

different_position_results$Position <- factor(different_position_results$Position, 
                                              levels = c("word-initial", 
                                                         "word-medial", 
                                                         "word-final"))

ggplot(
  different_position_results,
  aes(x = LexifierPhoneme, y = m, fill = Position)
) +
  geom_col(position = position_dodge2(width = 0.9, preserve = "single")) +
  labs(x = "Lexifier phoneme", y = "Mean stability", fill = "Word position")
```

Flip horizontally.

```{r, warning=FALSE, message=FALSE}
ggplot(different_position_results) +
  geom_bar(
    aes(
      x = m,
      y = reorder(LexifierPhoneme, m),
      fill = Position
    ),
    stat = "identity",
    show.legend = TRUE,
    position = "dodge2"
  ) +
  labs(x = "Stability score", y = "Phoneme", fill = "Word position")
```

## Typological frequency and borrowability

First, we turn the data into ordinal values. Ordinal data was generated by ranking the percentage values of stability, borrowability and typological frequency from 1 to 19. Duplicate values summed and averaged in the ranking.

Cross linguistic frequency

```{r, warning=FALSE, message=FALSE}

Typology <- c(1, 2, 3, 4, 5, 5, 7, 8, 9, 10, 11.5, 11.5, 13, 14, 15, 16, 17, 18, 19)
consonant <- c("m", "k", "p", "n", "t", "l", "s", "b", "g", "d", "f", "r", "ɲ", 
               "t̠ʃ","z","v","ɾ","ʒ","ʎ")
df_typ <- data.frame(Typology, consonant)
```

2) Borrowability

```{r, warning=FALSE, message=FALSE}
Borrowability <- c(1, 2, 3, 4, 5, 6, 6, 8, 9, 10, 11, 12, 13, 14.5, 14.5, 16, 17, 18, 19)
consonant <- c("f", "g", "t̠ʃ","b","z","v","d","r","p","l","s","ʒ","ɾ","ɲ","k"
               ,"ʎ","t","n","m")
df_bor <- data.frame(Borrowability, consonant)
```

3) Stability values

```{r, warning=FALSE, message=FALSE}
Stability <- c(2, 2, 2, 2, 5, 7, 7, 7, 8, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19)
consonant <- c("t", "p", "n", "m", "f", "b", "k", "g", "d", "z", "s", "l", "r", 
               "t̠ʃ" ,"ɾ"  ,"ɲ" ,"ʒ" ,"ʎ" ,"v")
df_sta <- data.frame(Stability, consonant)

```

Then, we create the data frames and prepare them for non-parametric tests. Long format with joint groups.

```{r, warning=FALSE, message=FALSE}

df_friedman <- left_join(df_sta, df_bor, by="consonant")

order_df <- left_join(df_friedman, df_typ, by="consonant")
head(order_df)

df_long <- order_df %>% 
  gather(key = "conditions",value = "order", Borrowability, Stability, Typology)

head(df_long)
```

For the Spearman's rank correlation coefficient, we use large format and separated groups.

```{r, warning=FALSE, message=FALSE}
df_sta_bor <- left_join(df_sta, df_bor, by="consonant")
df_sta_typ <- left_join(df_sta, df_typ, by="consonant")

```

Convert to long format.

```{r, warning=FALSE, message=FALSE}
  
sta_bor_long <- df_sta_bor %>% 
  gather(key = "conditions", value = "order", Borrowability, Stability)

sta_typ_long <- df_sta_typ %>% 
  gather(key = "conditions", value = "order", Typology, Stability)
```

Finally, we perform the non-parametric tests. The statistical summary:

```{r, warning=FALSE, message=FALSE}
df_long %>% group_by(conditions) %>%  
  summarise(n = n(), mean = mean(order), sd = sd(order))
```

A first plot.

```{r, warning=FALSE, message=FALSE}
ggplot(df_long, aes(x = consonant, y = order)) + 
  geom_boxplot(outlier.shape = NA) + 
  geom_jitter(width = 0.2) + 
  theme(legend.position="top") +
  ylab("Order") +
  xlab("Consonant") # @CS: FIX
```

1. Friedman test.

```{r, warning=FALSE, message=FALSE}

friedman.test(y = df_long$order, groups = df_long$conditions, blocks = df_long$consonant)

df_long %>% friedman_effsize(order ~ conditions | consonant)
```

2. Conover's all-pairs test.

```{r, warning=FALSE, message=FALSE}

frdAllPairsConoverTest(
  y = df_long$order, 
  groups = df_long$conditions, 
  blocks = df_long$consonant, 
  p.adjust.method = "bonf")
```

3. Durbin's all-pairs test.

```{r, warning=FALSE, message=FALSE}
durbinAllPairsTest(
  y      = df_long$order, 
  groups = df_long$conditions, 
  blocks = df_long$consonant,
  p.adjust.method = "holm")
```

4. Spearman's correlation coefficient.

4.1 Stability ~ Borrowability.

```{r, warning=FALSE, message=FALSE}
cor.test(x=df_sta_bor$Borrowability, 
         y=df_sta_bor$Stability, 
         method = 'spearman')
```

4.2 Stability ~ Typological frequency.

```{r, warning=FALSE, message=FALSE}
cor.test(x=df_sta_typ$Typology, 
         y=df_sta_typ$Stability, 
         method = 'spearman')
```

Next, visualize the results. Box plots.

Stability vs typological frequency.

```{r, warning=FALSE, message=FALSE}
ggplot(sta_typ_long, aes(x = consonant, y = order)) + geom_boxplot(outlier.shape = NA) + 
  geom_jitter(width = 0.2) + theme(legend.position="top") +
  ylab("Order") +
  xlab("Consonant") # @CS: FIX

```

Stability vs borrowability.

```{r, warning=FALSE, message=FALSE}
ggplot(sta_bor_long, aes(x = consonant, y = order)) + geom_boxplot(outlier.shape = NA) + 
  geom_jitter(width = 0.2) + theme(legend.position="top") +
  ylab("Order") +
  xlab("Consonant") # @CS: FIX
```

Bump chart.

```{r, universal_tendencies, warning=FALSE, message=FALSE}

ggplot(data = df_long, aes(x = conditions, y = order, group = consonant)) +
  geom_line(aes(color = consonant, alpha = 1), size = 1) +
  geom_point(aes(color = consonant, alpha = 1), size = 2,  alpha = 0.8) +
  scale_y_reverse(breaks = 1:nrow(df_long)) +
  scale_alpha(guide = 'none') +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks.x = element_blank())+
  ylab("Ranking")+
  xlab("")
```


## Inventory size and frequency across substrates

Get the data.

```{r}
inv <- read.csv("Inventories.csv")
head(inv)
```

Prepare the data.

```{r}
df_inv_long <- inv %>% dplyr::select(Language, Phoneme) %>%  mutate(newcol = 1)

df_inv <- df_inv_long %>% pivot_wider(names_from = Language, 
                                      values_from = newcol, values_fill = 0)

df_total_inv <- df_inv %>% mutate_at(c(2:38), as.numeric)

head(df_total_inv)
```

Measuring the inventory size. Get the consonant inventory size for all languages.

```{r, warning=FALSE, message=FALSE}
cons_count <- df_total_inv %>% dplyr::select(c(2:38)) %>% mutate_at(c(1:37), as.numeric)
  
count <- colSums(cons_count [,c(1:37)]) #%>% unname(colSums(count))

cons_lg <- dplyr::select(df_inv_long, "Language") 

Language <- unique(cons_lg$Language)

category <- inv %>% dplyr::select(Language, Category)

count_lg <- data.frame(cbind(Language, count)) 
  
count_lg_1 <- inner_join(count_lg, category, by = "Language") %>% distinct()

inv_size <- transform(count_lg_1, count = as.numeric(count))

head(inv_size)

```

Which languages have bigger inventories?

```{r}
ggplot(inv_size) + 
  geom_bar(aes(x = Language,
  y = count,
  fill = Category), 
  stat = "identity", 
  show.legend = TRUE,
  position = "dodge2") + coord_flip() +
  xlab("Language") +
  ylab("Count") # @CS: FIX
```

Violin plot shows that the majority of creoles have larger consonant inventories than Portuguese.

```{r, size_violin, warning=FALSE, message=FALSE}
ggplot(inv_size, aes(x = Category, y = count, fill = Category)) +
  geom_violin(alpha = 0.5) +
  geom_dotplot(binaxis = "y",
               stackdir = "center",
               dotsize = 0.5) +
  theme(legend.position = "none") +
  ylab("Number of consonants")
```

Consonant frequency in all languages involved. Count frequent consonants.

```{r}
total <- rowSums(cons_count)

cons_freq <- data.frame(cbind(df_total_inv$Phoneme, total))

cons_freq <- transform(cons_freq, total = as.numeric(total))

colnames(cons_freq)[1] <- "LexifierPhoneme"

```

Is there a relationship between this frequency and the stability values? Portuguese consonants only.

```{r}
cons_freq_pt <- cons_freq %>% subset(LexifierPhoneme %in% 
                                       c('b','d','f','g','k','l','ʎ','m','n',
                                         'ɲ','p','t̠ʃ','ɾ','s','t','v','z','ʒ', 'r'))

```

First dataset: relative frequency values.

```{r}
cons_freq_rel <- cons_freq_pt %>% mutate(frequency = total/37)
```

Second dataset: stability values.

```{r}
consonant_global_stability <- read.csv("consonant_global_stability.csv")
```

Merge the datasets.

```{r}
cor_freq_sta <- left_join(consonant_global_stability, cons_freq_rel, by='LexifierPhoneme')
```

Results of a regression analysis.

```{r}
fs <- lm(frequency ~ mglobal, data=cor_freq_sta)
summary(fs)
```

Plot the results.

```{r, warning=FALSE, message=FALSE}

fs_plot <- ggplot(cor_freq_sta, aes(x = frequency, y = mglobal, label = LexifierPhoneme)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() + 
  ylab("Mean global score") +  # @CS: FIX
  xlab("Frequency") # +
  #geom_text(aes(label=V1), hjust=3, vjust=0)

fs_plot + geom_text_repel(aes(label=LexifierPhoneme))
```

There is a relationship between stability and frequency across all languages in the sample. But does it make sense? We are measuring the consonants in all categories and the present of the lexifier and the creoles may influence the results. Thus, we subset and measure the frequency across substrates only, but this procedure may just increase the lack of correlation.

Is there a relationship between consonant stability and their presence in the inventories of the substrate languages?

Data preparation (substrates only).

```{r}
inv_subs <- inv %>% subset(Category == 'substrate') %>% 
  dplyr::select(Language, Phoneme) %>% 
  mutate(newcol = 1)

inv_subs_long <- inv_subs %>% 
  pivot_wider(names_from = Language, values_from = newcol, values_fill = 0)

inv_subs_long <- inv_subs_long %>% 
  mutate_at(c(2:18), as.numeric)

head(inv_subs_long)

```

Sum row values and subset to consonants which have correspondents in Portuguese.

```{r}
subs_count <- inv_subs_long %>% 
  dplyr::select(c(2:18))

total_subs <- rowSums(subs_count)

subs_freq <- data.frame(cbind(inv_subs_long$Phoneme, total_subs))

subs_freq <- transform(subs_freq, total_subs = as.numeric(total_subs))

colnames(subs_freq)[1] <- "LexifierPhoneme"

subs_freq <- subs_freq %>% 
  subset(LexifierPhoneme %in% c('b','d','f','g','k','l','ʎ','m','n','ɲ','p',
                                't̠ʃ','ɾ','s', 't','v','z','ʒ', 'r'))

```

Get relative values.

```{r}
subs_freq_rel <- subs_freq %>% mutate(frequency = total_subs/17)
```

Merge datasets.

```{r}
subs_sta <- left_join(consonant_global_stability, subs_freq_rel, by='LexifierPhoneme')
```

Results of a simple regression.

```{r}
subs_sta_lm <- lm(frequency ~ mglobal, data=subs_sta)
summary(subs_sta_lm)
```

Plot the results.

```{r, substratefreq_creosta, warning=FALSE, message=FALSE}
subsfreq_sta_cor <- ggplot(subs_sta, aes(x = frequency, y = mglobal, label = LexifierPhoneme)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  xlab("Frequency across substrates") + ylab("Stability across creoles") # +
 # geom_text(aes(label=V1), hjust=3, vjust=0)


subsfreq_sta_cor + geom_text_repel(aes(label=LexifierPhoneme))
```

When we consider the correlation between the consonant stability in creoles and the frequency of these consonants in the substrates only, we find that there is a weaker correlation (if we compare with the results above). However, this correlation is statistically significant (p-value: 0.004). Nothing the outliers which are normally voiced consonants.

Typological frequency vs. Substrate frequency.

Typological frequency data, extrated from PHOIBLE.

```{r}
typ_freq <- read.csv("typ_freq.csv")
```

Merge the datasets.

```{r}
#consonant_global_stability <- read.csv("consonant_global_stability.csv")
typ_sta <- left_join(typ_freq, consonant_global_stability, by='LexifierPhoneme')
```

Results of a linear regression.

```{r}
typ_sta_lm <- lm(TypologicalFreq ~ mglobal, data=typ_sta)
summary(typ_sta_lm)
```

Plot the results.

```{r, typfreq_stacreo, warning=FALSE, message=FALSE}
typ_sta_cor <- ggplot(typ_sta, aes(x = TypologicalFreq, y = mglobal, label = LexifierPhoneme)) +
  geom_smooth(method = "lm", colour="black") +
  geom_point() +
  xlab("Frequency across the world's languages") + ylab("Stability across creoles") # +
 # geom_text(aes(label=V1), hjust=3, vjust=0)

typ_sta_cor + geom_text_repel(aes(label=LexifierPhoneme))
```


# References
